---
title: "Kang dataset"
author: "Almut Lütge"
date: "2 März 2020"
output:
  html_document:
    self_contained: no
    lib_dir: '../docs/site_libs'
    code_folding: show
    theme: journal
    highlight: tango
    number_sections: no
    toc: yes
    toc_depth: 3
    toc_float: true
    collapsed: no
    smooth_scroll: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Kang dataset
10x droplet-based scRNA-seq PBMC data from 8 Lupus patients before and after 6h-treatment with INF-beta (16 samples in total). data are derived from the muscData package and can be assessed via ExperimentHub. Here we use only the control sample to not mix treatment and batch effect.
See https://bioconductor.org/packages/release/data/experiment/vignettes/muscData/inst/doc/muscData.html.


```{r libraries}
suppressPackageStartupMessages({
    library(plotly)
    library(readr)
    library(stringr)
    library(edgeR)
    library(stringr)
    library(pheatmap)
    library(purrr)
    library(scater)
    library(dplyr)
    library(reshape2)
    library(ggplot2)
    library(cowplot)
    library(Matrix)
    library(scran)
    library(LSD)
    library(Seurat)
    library(sctransform)
    library(readxl)
    library(DropletUtils)
    library(CellMixS)
    library(tibble)
    library(ExperimentHub)
})

seed <- 1000
```

# Data
Load data
```{r data}
out_path <- here::here("out")
sc <- ExperimentHub()
sce <- sc[["EH2259"]]
## Filter out genes that are not expressed in any cell
sce <- sce[which(rowSums(counts(sce) > 0) > 0), ]
sce$patient <- sce$ind

sce$patient <- factor(sce$patient)

table(sce$patient)
table(sce$patient, sce$stim)
dim(sce)
dim(colData(sce))
dim(rowData(sce))

#remove stmulated sample
sce <- sce[,!sce$stim %in% "stim"]
dim(sce)

#remove doublets
sce <- sce[,!sce$multiplets %in% "doublet"]
dim(sce)
```

### Calculate QC
```{r qc}
#remove genes without any counts
keep_features <- rowSums(counts(sce) > 0) > 0
sce <- sce[keep_features, ]
dim(sce)

# # Mitochondrial genes
is.mito <- grepl("MT-", rownames(sce))
summary(is.mito)
mito <- rownames(sce)[is.mito]

sce <- calculateQCMetrics(sce, feature_controls = list(Mt = mito))

```

# Filtering 
## Find outlier
```{r filter cells, fig.height = 10, fig.width = 12}
# # Plot filters
plotFilters <- function( sce, var="log10_total_counts", split_by="patient", nrow=NULL,
                         nmads=c(2,3,5), lt=c("dashed","dotted","dotdash"), xscale="free" ){
  CD <- as.data.frame(colData(sce))
  if(!(var %in% colnames(CD))) stop(paste("`var`",var,"is not in `colData(sce)`!"))
  if(!is.null(split_by) && !(split_by %in% colnames(CD))){
    stop(paste("`split_by`",split_by,"is not in `colData(sce)`!"))
  }
  library(ggplot2)
  library(cowplot)
  d <- CD[,var,drop=F]
  if(!is.null(split_by)) d$dataset <- CD[[split_by]]
  p <- ggplot(d, aes_string(x=var)) + geom_histogram(color="darkblue", bins=30)
  if(xscale!="free"){
    if(xscale!="fixed"){
      if(xscale>1 && xscale%%1==0){
        xq <- .tmads(d[[var]], xscale)
        xr <- range(d[[var]],na.rm=T)
        xq <- c(max(xq[1],xr[1]), min(xq[2],xr[2]))
      }else{
        if(xscale<=1 & xscale>0){
          xscale <- (1-xscale)/2
          xq <- quantile(d[[var]], probs=c(xscale,1-xscale), na.rm=T)
        }else{
          stop("Wrong `xscale` value!")
        }
      }
      p <- p + xlim(xq[1], xq[2])
    }
  }
  if(!is.null(split_by)){
    if(is.null(nrow)) nrow <- ceiling(length(unique(d$dataset))/3)
    p <- p + facet_wrap(~dataset, scales=ifelse(xscale=="free","free","free_y"), nrow=nrow)
    for(ds in unique(d$dataset)){
      for(i in 1:length(nmads)){
        ma <- .tmads(d[which(d$dataset==ds),var], nmads[i])
        df2 <- data.frame(xint=as.numeric(ma), dataset=rep(ds,2))
        p <- p + geom_vline(data=df2, aes(xintercept=xint), linetype=lt[i])
      }
    }
  }else{
    for(i in 1:length(nmads)){
      df2 <- data.frame(xint=as.numeric(.tmads(d[[var]], nmads[i])))
      p <- p + geom_vline(data=df2, aes(xintercept=xint), linetype=lt[i])
    }
  }
  p
}
.tmads <- function(x, nbmads=2.5){
  x2 <- nbmads*median(abs(x-median(x)))
  median(x)+c(-x2,x2)
}
plotFilters(sce)
plotFilters(sce, "log10_total_features_by_counts")
plotFilters(sce, "pct_counts_Mt", xscale=0.98)

# Find outlier
outlierPlot <- function(cd, feature, aph=NULL, logScale=FALSE, show.legend=TRUE){
  if(is.null(aph)) aph <- paste0(feature, "_drop")
  if(!(aph %in% colnames(cd))) aph <- NULL
  p <-  ggplot(as.data.frame(cd), aes_string(x = feature, alpha = aph)) +
    geom_histogram(show.legend=show.legend)
  if(!is.null(aph)) p <- p + scale_alpha_manual(values = c("TRUE" = 0.4, "FALSE" = 1))
  if(logScale) p <- p + scale_x_log10()
  p
}
plQCplot <- function(cd, show.legend=TRUE){
  ps <- lapply(split(cd,cd$patient), sl=show.legend, FUN=function(x,sl){
    list( outlierPlot( x, "total_counts", logScale=T, show.legend=sl),
          outlierPlot( x, "total_features_by_counts", "total_features_drop",
                       logScale=T, show.legend=sl),
          outlierPlot( x, "pct_counts_Mt", "mito_drop", show.legend=sl)
    )
  })
  plot_grid( plotlist = do.call(c, ps),
             labels=rep(basename(names(ps)), each=length(ps[[1]])),
             ncol=length(ps[[1]]),
             label_x=0.5 )
}


#Filtering
sce$total_counts_drop <- isOutlier(sce$total_counts, nmads = 2.5,
                                   type = "both", log = TRUE, batch=sce$patient)
sce$total_features_drop <- isOutlier(sce$total_features_by_counts, nmads = 2.5,
                                     type = "both", log = TRUE, batch=sce$patient)
sce$mito_drop <- sce$pct_counts_Mt > 5 &
  isOutlier(sce$pct_counts_Mt, nmads = 2.5, type = "higher", batch=sce$patient)

sce$isOutlier <- sce$total_counts_drop | sce$total_features_drop | sce$mito_drop

# quality plot
plQCplot(colData(sce), show.legend=FALSE)

ggplot(colData(sce) %>% as.data.frame, aes(x=total_features_by_counts, y=total_counts, colour=pct_counts_Mt)) + geom_point() + facet_wrap(~patient) +geom_density_2d(col="white") + scale_x_sqrt() + scale_y_sqrt()

ggplot(colData(sce) %>% as.data.frame, aes(x=total_features_by_counts, y=pct_counts_Mt)) + geom_point() + facet_wrap(~patient) +geom_density_2d(col="white")

```

## Check thresholds
```{r threshholds, fig.width = 12, fig.height = 5}
# Check outlier
mets <- c("total_counts_drop","total_features_drop","mito_drop")
sapply(mets, FUN=function(x){ sapply(mets, y=x, function(x,y){ sum(sce[[x]] & sce[[y]]) }) })
nbcells <- cbind(table(sce$patient),table(sce$patient[!sce$isOutlier]))
colnames(nbcells) <- c("cells total","cells after filtering")
nbcells
layout(matrix(1:2,nrow=1))
LSD::heatscatter( sce$total_counts, sce$total_features_by_counts, xlab="Total counts", ylab="Non-zero features", main="",log="xy")
w <- which(!sce$isOutlier)
LSD::heatscatter( sce$total_counts[w], sce$total_features_by_counts[w], xlab="Total counts", ylab="Non-zero features", main="Filtered cells",log="xy")
# summary of cells kept
cct <- table(sce$isOutlier, sce$patient)
row.names(cct) <- c("Kept", "Filtered out")
cct
# drop outlier cells
sce <- sce[,!sce$isOutlier]

# require count > 1 in at least 20 cells
sce <- sce[which(rowSums(counts(sce)>1)>=20),]
dim(sce)
plQCplot(colData(sce), show.legend=FALSE)

table(sce$patient)

```


# Save data
```{r saving}
# Save data
saveRDS(sce, file = paste0(out_path, "/sce_kang.rds"))
```

```{r session info}
sessionInfo()
```

