---
title: "pbmc_roche"
author: "Almut Lütge"
date: "14 March 2021"
output:
  html_document:
    self_contained: no
    lib_dir: '../docs/site_libs'
    code_folding: show
    theme: journal
    highlight: tango
    number_sections: no
    toc: yes
    toc_depth: 3
    toc_float: true
    collapsed: no
    smooth_scroll: yes
    dev: 'png'
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Pbmc dataset Roche
This datset has been prepared by Roche. It contains pbmc from 4 different sample. 
Sample are derived from the same patient, have been processed in the same way and have been sequenced together. Experimental differences are due to different storage conditions. One sample was sequenced from fresh cells, the other ones have been stored in different media and frozen for a week.


```{r libraries}
suppressPackageStartupMessages({
  library(pheatmap)
  library(purrr)
  library(scater)
  library(dplyr)
  library(ggplot2)
  library(cowplot)
  library(scran)
  library(CellMixS)
  library(magrittr)
  library(here)
  library(SingleR)
  library(celldex)
  library(Seurat)
  library(broom)
  library(tidyr)
  library(ggridges)
})

seed <- 1000
```


### Data
Load data
For processing check datasets: [pbmc_roche](https://almutlue.github.io/batch_dataset/pbmc_roche.html)
```{r data}
data_path <- here::here("out")
sce <- readRDS(file = paste0(data_path, "/sce_pbmc_roche.rds"))

dim(sce)

#show data
visGroup(sce, group = "batch", dim_red = "tsne")
visGroup(sce, group = "cluster", dim_red = "tsne")

```

# Celltype assignment singleR 
```{r singleR, warning=FALSE, message=FALSE}
ref <- BlueprintEncodeData()
rowData(sce)$ENS_ID <- gsub("\\..*","", rownames(sce))
rownames(sce) <- gsub(".*\\.","", rownames(sce))
pred <- SingleR(test=sce, ref=ref, labels=ref$label.main)
table(pred$labels)

plotScoreHeatmap(pred)
```


### Compare singleR assignments with clustering
```{r control cluster, warning = FALSE}
tab <- table(Assigned=pred$pruned.labels, Cluster=sce$cluster)
pheatmap(log2(tab+10), color=colorRampPalette(c("white", "blue"))(101))
sce$singleR <- pred$pruned.labels
visGroup(sce, group="singleR", dim_red = "tsne")
```


## Check marker genes

### Marker genes for "seurat clustering" {.tabset}

```{r marker seurat, fig.width = 5, fig.height=7, warning = FALSE, results = "asis"}
marker_s <- findMarkers(sce, sce$cluster)

#plot summarized counts
plot_aggregated_expr <- function(sce, genes, group_var, title){
  logNormExpres <- as.data.frame(t(as.matrix(logcounts(sce)[genes,]))) %>% 
    dplyr::mutate(cluster_all= colData(sce)[,group_var]) %>%
    group_by(cluster_all) %>% summarise_all(mean)
  logNormExpresMa <- logNormExpres %>% set_rownames(logNormExpres$cluster_all) %>% 
    dplyr::select(-cluster_all) %>% as.matrix() %>% t()
  colnames(logNormExpresMa) <- levels(as.factor(colData(sce)[,group_var]))
  rownames(logNormExpresMa) <- rownames(logNormExpresMa)
  p <- pheatmap(logNormExpresMa, scale="row" ,treeheight_row = 0, 
           cluster_cols = F, cluster_rows = F, 
           color = colorRampPalette(c("#2166AC", "#F7F7F7", "#B2182B"))(50), 
           main=title, cellwidth=15, cellheight=10)
}


#plot marker
clust_names <- names(marker_s)

for (clust in clust_names) {
  cat("#### ", clust, "{-}\n")
  cluster_res <- marker_s[[clust]]
  top_marker <- cluster_res[cluster_res$Top <= 6,]
  print(p <- plot_aggregated_expr(sce, rownames(top_marker), 
                      group_var = "cluster", 
                      title = paste0("pbmc_seurat_", clust)))
  cat("\n\n")
}

```


### Marker genes for "SingleR annotations" {.tabset}
```{r marker singelR, fig.width = 4, fig.height=6, results = "asis", warning = FALSE}
#filter cell types with less than 10 cells
ct_filter <- names(which(table(sce$singleR) < 10))
sce <- sce[,!sce$singleR %in% ct_filter]
sce <- sce[,which(!is.na(sce$singleR))]

marker_single <- findMarkers(sce, sce$singleR)

#plot marker
clust_names <- names(marker_single)

for (clust in clust_names) {
  cat("#### ", clust, "{-}\n")
  cluster_res <- marker_single[[clust]]
  top_marker <- cluster_res[cluster_res$Top <= 6,]
  print(p <- plot_aggregated_expr(sce, rownames(top_marker), 
                      group_var = "singleR", 
                      title = paste0("pbmc_singleR_", clust)))
  
  cat("\n\n")
}


```

# DE analysis

## Wilcoxon and t-test 

### DE genes between all cells
```{r de wilcox all}
# All conditions , blocking for celltypes
de_wilcox <- findMarkers(sce, groups = sce$batch, 
                      test.type = "wilcox", block = sce$singleR,
                      full.stats = TRUE, log.p = TRUE)

de_ttest <- findMarkers(sce, groups = sce$batch, test.type= "t", block = sce$singleR,
                    full.stats = TRUE)

summarize_de <- function(de_list, target){
  all_res <- lapply(names(de_list), function(cond){
    cond2_list <- names(de_list)[!names(de_list) %in% cond]
    cond_res <- lapply(cond2_list, function(cond2){
      de_all <- de_list[[cond]][[paste0("stats.", cond2)]]
      if( target %in% "AUC"){
        de_filterted <- de_all[abs(de_all$AUC - 0.5) > 0.05 & exp(de_all$log.FDR) < 0.01,]
      }else{
        de_filterted <- de_all[ exp(de_all$log.FDR) < 0.01,]
      }
      n_de <- nrow(de_filterted)
    }) %>% unlist() %>% cbind(.,cond2_list) %>% as.data.frame() %>% set_colnames(c("DE", "batch2"))
  }) %>% set_names(names(de_list)) %>% bind_rows(.id = "batch1")
}

summarize_dist <- function(de_list, target){
  all_res <- lapply(names(de_list), function(cond){
    cond2_list <- names(de_list)[!names(de_list) %in% cond]
    cond_res <- lapply(cond2_list, function(cond2){
      de_all <- de_list[[cond]][[paste0("stats.", cond2)]]
      if( target %in% "AUC"){
        dist <- de_all$AUC
      }else{
        dist <- de_all$logFC
      }
      dist
    }) %>% set_names(cond2_list)  %>% bind_cols()
  }) %>% set_names(names(de_list))
}

sum_de_wilcox <- summarize_de(de_wilcox, target = "AUC")
sum_de_t <- summarize_de(de_ttest, target = "logFC")
sum_dist_wilcox <- summarize_dist(de_wilcox, target = "AUC")
sum_dist_t <- summarize_dist(de_ttest, target = "logFC")


```

### Plot DE {.tabset}

#### Wlicoxon
```{r plot DE wilcox}

sum_de_wilcox$DE <- as.numeric(sum_de_wilcox$DE)

# Plot de_genes
p <- ggplot(data = sum_de_wilcox, aes(x = batch1, y = batch2, colour= DE)) +
      geom_point(aes(size = DE)) +
      theme(axis.title.x = element_blank(), axis.title.y = element_blank()) +
      scale_colour_viridis_c() +
      labs(
        x = 'Batch',
        y = 'Batch') +
    guides(color= guide_legend(), size=guide_legend()) +
      theme_cowplot() +
    scale_size(range = c(4,14)) +
    ggtitle("Filtered DE genes using Wilcoxon test")

p 


```

#### t-test

```{r de ttest}
sum_de_t$DE <- as.numeric(sum_de_t$DE)

# Plot de_genes
p <- ggplot(data = sum_de_t, aes(x = batch1, y = batch2, colour= DE)) +
      geom_point(aes(size = DE)) +
      theme(axis.title.x = element_blank(), axis.title.y = element_blank()) +
      scale_colour_viridis_c() +
      labs(
        x = 'Batch',
        y = 'Batch') +
    guides(color= guide_legend(), size=guide_legend()) +
      theme_cowplot() +
    scale_size(range = c(4,14)) +
    ggtitle("DE t-test")

p 
```

### Plot logFC distributions {.tabset}

```{r ggranges, results="asis"}

plot_dist <- function(batch_nam, tab){
  tab_dist <- tab[[batch_nam]]
  gathercols <- colnames(tab_dist)
  dist_long <- gather(tab_dist, "batch", "valuecol", all_of(gathercols),
                           factor_key=TRUE)
  ggplot(dist_long, aes_string(x="batch", y="valuecol", fill="batch")) +
            geom_violin()  +
            labs(title=paste0("Expression differences compared to ", batch_nam),
                 x="batch",
                 y = "logFC") +
            scale_fill_brewer(palette="Dark2") + 
  theme_cowplot()
}

for (batch in names(sum_dist_t)) {
  cat("#### ", batch, "{-}\n")
  ps <- plot_dist(batch, tab = sum_dist_t)
  print(ps)
  cat("\n\n")
}


```


### Cell-type-wise DE {.tabset}

```{r ct de, results="asis"}

for (ct in levels(as.factor(sce$singleR))) {
  cat("#### ", ct, "{-}\n")
  sce_new <- sce[, sce$singleR %in% ct]
  de_wilcox <- findMarkers(sce_new, groups = sce_new$batch, 
                      test.type = "wilcox",
                      full.stats = TRUE, log.p = TRUE)
  sum_de_wilcox <- summarize_de(de_wilcox, target = "AUC")
  sum_de_wilcox$DE <- as.numeric(sum_de_wilcox$DE)
  p <- ggplot(data = sum_de_wilcox, aes(x = batch1, y = batch2, colour= DE)) +
      geom_point(aes(size = DE)) +
      theme(axis.title.x = element_blank(), axis.title.y = element_blank()) +
      scale_colour_viridis_c() +
      labs(
        x = 'Batch',
        y = 'Batch') +
      guides(color= guide_legend(), size=guide_legend()) +
      theme_cowplot() +
      scale_size(range = c(4,14)) +
      ggtitle(paste0("filtered DE Wilcoxon test in ", ct))
  print(p)
  cat("\n\n")
}

```
### session Info
```{r session info}
sessionInfo()
```

